---
title: "Assignment 4: Data Wrangling"
author: Cristiana Falvo
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 4 at 1:00 pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1 
getwd()
library(tidyverse)
library(lubridate)
# data
EPA_Ozone_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv")
EPA_Ozone_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv")
EPA_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
EPA_PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv")

#2
## explore ozone data
head(EPA_Ozone_2018)
head(EPA_Ozone_2019)
dim(EPA_Ozone_2018)
dim(EPA_Ozone_2019)
str(EPA_Ozone_2018)
str(EPA_Ozone_2019)

# are column names the same? 
ozone_18_names <- colnames(EPA_Ozone_2018)
ozone_19_names <- colnames(EPA_Ozone_2019)
class(ozone_18_names)

intersect(ozone_18_names, ozone_19_names)
setdiff(ozone_18_names, ozone_19_names)
setdiff(ozone_19_names, ozone_18_names)
# ..yes, column names are identical

## explore particulate matter (PM) data
head(EPA_PM25_2018)
head(EPA_PM25_2019)
dim(EPA_PM25_2018)
dim(EPA_PM25_2019)
str(EPA_PM25_2018)
str(EPA_PM25_2019)

# compare PM column names
PM_18_names <- colnames(EPA_PM25_2018)
PM_19_names <- colnames(EPA_PM25_2019)
intersect(PM_18_names, PM_19_names)
# pm names also match from 2018 to 2019

# how about PM vs ozone?
intersect(ozone_18_names, PM_18_names)
setdiff(ozone_18_names, PM_18_names)
setdiff(PM_18_names, ozone_18_names)
# one field is different between ozone and PM data, the rest are the same
```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 
# change date for ozone 2018 data
EPA_Ozone_2018$Date <- as.Date(EPA_Ozone_2018$Date, format = "%m/%d/%Y")
class(EPA_Ozone_2018$Date)
# change date for ozone 2019 data
EPA_Ozone_2019$Date <- as.Date(EPA_Ozone_2019$Date, format = "%m/%d/%Y")
class(EPA_Ozone_2019$Date)
# change date for PM 2018 data
EPA_PM25_2018$Date <- as.Date(EPA_PM25_2018$Date, format = "%m/%d/%Y")
class(EPA_PM25_2018$Date)
# change date for PM 2019 data
EPA_PM25_2019$Date <- as.Date(EPA_PM25_2019$Date, format = "%m/%d/%Y")
class(EPA_PM25_2019$Date)

#4
# select columns for ozone 2018
ozone_2018_subset <- select(EPA_Ozone_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
# select columns for ozone 2019
ozone_2019_subset <- select(EPA_Ozone_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
# select columns for PM 2018
PM_2018_subset <- select(EPA_PM25_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
# select columns for PM 2019
PM_2019_subset <- select(EPA_PM25_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

class(ozone_2018_subset$Date)
class(ozone_2019_subset$Date)


#5
# change desc field of PM 2018 subset data
PM_2018_subset_adjusted <-
  PM_2018_subset %>%
  mutate(AQS_PARAMETER_DESC = "PM2.5")
# change desc field of PM 2019 subset data
PM_2019_subset_adjusted <-
  PM_2019_subset %>%
  mutate(AQS_PARAMETER_DESC = "PM2.5")

class(PM_2018_subset_adjusted$Date)
class(PM_2019_subset_adjusted$Date)

#6
# save ozone 2018 processed data
write.csv(ozone_2018_subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
# save ozone 2019 processed data
write.csv(ozone_2019_subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
# save PM 2018 processed data
write.csv(PM_2018_subset_adjusted, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
# save PM 2019 processed data
write.csv(PM_2019_subset_adjusted, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
# I want to see unique values in the site column
unique(ozone_2018_subset$Site.Name)
unique(ozone_2019_subset$Site.Name)
unique(PM_2018_subset_adjusted$Site.Name)
unique(PM_2019_subset_adjusted$Site.Name)
# dimensions of four sep datasets
dim(ozone_2018_subset)
dim(ozone_2019_subset)
dim(PM_2018_subset_adjusted)
dim(PM_2019_subset_adjusted)

#7 
# combined four datasets into one
combined_ozone_PM <- rbind(ozone_2018_subset, ozone_2019_subset, PM_2018_subset_adjusted, PM_2019_subset_adjusted)
dim(combined_ozone_PM) # 37,893 rows

combined_ozone_PM$Date <- as.Date(combined_ozone_PM$Date, format = "%Y-%m-%d")
class(combined_ozone_PM$Date)


#8
combined_sites <-
  combined_ozone_PM %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School"))

# seeing the difference between full combo and combo w/ sp sites
## str and sitename outputs say same amount of sites (51 levels, # of sites in full combo)
## I know that R is just storing all the sites behind the scenes, and the combo_sites is only specified sites after testing..
### showing intersect between full and sites combo shows the correct/expected commonalities
### viewing combo_sites dataframe, there was no data for sites we didn't select in filter function
intersect(combined_ozone_PM$Site.Name, combined_sites$Site.Name)
str(combined_ozone_PM)
str(combined_sites)
combined_ozone_PM$Site.Name
combined_sites$Site.Name

# trying to look piece by piece at group_by function to see what it's doing..see notebook for Q
combined_group_date_desc <- 
  combined_sites %>%
  group_by(Date, AQS_PARAMETER_DESC) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE))

# view(combined_group_date)
# view(combined_group_date_desc)

combined_narrowed <-
  combined_ozone_PM %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE)) %>%
  mutate(month = month(Date), year = year(Date)) %>%
  select(month, year, Date:meanLONG)
  
dim(combined_narrowed) #14,752 x 9 -- as expected

# 9  Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
combined_processed <- spread(combined_narrowed, AQS_PARAMETER_DESC, meanAQI)

#10
dim(combined_processed)

#11
write.csv(combined_processed, "..EPAair_O3_PM25_NC1718_Processed.csv")
```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
combined_processed_csv <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/EPAair_O3_PM25_NC1718_Processed.csv")

str(combined_processed_csv)

#12a
combined_summary <-
  combined_processed_csv %>%
  group_by(Site.Name, month, year) %>%
  summarise(meanAQI_ozone = mean(Ozone),
            meanAQI_pm25 = mean(PM2.5))

#12b remove instances where month and year are not available (drop_na)
combined_summary_na_drop <-
  combined_processed_csv %>%
  group_by(Site.Name, month, year) %>%
  summarise(meanAQI_ozone = mean(Ozone),
            meanAQI_pm25 = mean(PM2.5)) %>%
  drop_na()

combined_summary_na_omit <-
  combined_processed_csv %>%
  group_by(Site.Name, month, year) %>%
  summarise(meanAQI_ozone = mean(Ozone),
            meanAQI_pm25 = mean(PM2.5)) %>%
  na.omit()

#13
dim(combined_summary)


```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: After testing both they seem to do the same thing so perhaps it doesn't matter in this case.
