---
title: "Assignment 5: Data Visualization"
author: Cristiana Falvo
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 11 at 1:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse and cowplot packages. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (tidy and gathered) and the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
getwd()
library(tidyverse)
library(cowplot)
library(lubridate)
chem_nutrients <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
chem_physics <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
nutrients_wide <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")
nutrients_gathered <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
NIWOT <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")


#2
class(chem_nutrients$sampledate)
class(chem_physics$sampledate)
class(nutrients_gathered$sampledate)
class(nutrients_wide$sampledate)
class(NIWOT$collectDate)

chem_nutrients$sampledate <- as.Date(chem_nutrients$sampledate, format = "%Y/%m/%d")
chem_physics$sampledate <- as.Date(chem_physics$sampledate, format = "%Y-%m-%d")
nutrients_wide$sampledate <- as.Date(nutrients_wide$sampledate, format = "%Y/%m/%d")
nutrients_gathered$sampledate <- as.Date(nutrients_gathered$sampledate, format = "%Y/%m/%d")
NIWOT$collectDate <- as.Date(NIWOT$collectDate, format = "%Y/%m/%d")


```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#install.packages("devtools")
devtools::install_github("katiejolly/nationalparkcolors")
library(nationalparkcolors)

pal <- park_palette("SmokyMountains")

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") 

mytheme2 <- list(mytheme, scale_color_manual((values = pal)))

theme_set(mytheme) 
# when I did this and tried to make a plot it said many elements were missing
# went back to default for now

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus by phosphate, with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
Phosphorus_Phosphate <- 
  ggplot(nutrients_wide, aes(x = tp_ug, y = po4, color = lakename)) +
  geom_point() +
  xlim(0, 150) +
  ylim(0, 100) +
  geom_smooth(method = lm, color = "black", size = .5)
print(Phosphorus_Phosphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
# 5a
temp_box <- 
  ggplot(chem_nutrients) +
  geom_boxplot(aes(x = month, y = temperature_C, color = lakename))
print(temp_box)

# 5b
TP_box <-
  ggplot(chem_nutrients) +
  geom_boxplot(aes(x = month, y = tp_ug, color = lakename))
print(TP_box)

# 5c
TN_box <-
  ggplot(chem_nutrients) +
  geom_boxplot(aes(x = month, y = tn_ug, color = lakename))
print(TN_box)

# cowplot

all_box <- plot_grid(temp_box, TP_box, TN_box, ncol = 3, nrow = 1)
print(all_box)
  
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The nutrient concentrations are higher in Peter Lake than in Paul Lake. Change over time is hard to discern from these box plots..

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
class(NIWOT$collectDate)
S <- subset(NIWOT, functionalGroup == "Needles")

Needle_plot <- 
  ggplot(S, aes(x = as.factor(collectDate), y = dryMass, color = nlcdClass))  +
  geom_point() 
print(Needle_plot)

Needle_plot_facet <- 
  ggplot(S, aes(x = as.factor(collectDate), y = dryMass))  +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(Needle_plot_facet)

## don't know why collectDate can't plot on the x axis
  # without as.factor it gives error message
  # with as.factor it plots all dates as one (NA)



```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think looking at the plots separately in 3 facets is most effective because they are lined up for comparison but you can also see each individual dataset more clearly than when the colors are overlapping. 