---
title: "Assignment 7: GLMs week 2 (Linear Regression and beyond)"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 25 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory, load the tidyverse, nlme, and piecewiseSEM packages, import the *raw* NTL-LTER raw data file for chemistry/physics, and import the processed litter dataset. You will not work with dates, so no need to format your date columns this time.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
getwd()
library(tidyverse)
library(nlme)
library(piecewiseSEM)
# to visualize linear reg
require(ggiraph)
require(ggiraphExtra)
require(plyr)
# data
chem_phys <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
litter <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 
  
  # is this asking, what's the optimal combination of variables (aic..) to predict lake temps in july?

3. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years. (182-212)
* Only the columns: lakename, year4, daynum, depth, temperature_C (select())
* Only complete cases (i.e., remove NAs) (na.omit(data))

4. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 
  # run aic, this will give optimal combo of variables
  # with this optimal set of vars, run a 'multiple regression'

```{r}
#3 get subset of data
chem_phys_subset <- 
  chem_phys %>%
  filter(daynum >= 182 & daynum <= 212) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>% # 10,836
  na.omit(chem_phys_subset)

# see relationships between variables
  # hmmm what did we do in the gis lab..

#4 run AIC to determine best combination of predictor variables
temp_AIC <- lm(data = chem_phys_subset, temperature_C ~ year4 + daynum + depth) # told which eq to use
step(temp_AIC) # ..what does this tell us, the orig three is the best combo after all?
  # where it says "Call:" I think that's the recommended equation

temp_model <- lm(data = chem_phys_subset, temperature_C ~ year4 + daynum + depth)
summary(temp_model) 
# is this model the multiple regression?
# is there a way to visualize it?

# visualize individual relatinships (switch between year, daynum, depth - depth is only one with noteable correlation)
linear_reg_rel <- 
  ggplot(chem_phys_subset, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = .5) +
  geom_smooth(method = lm)
print(linear_reg_rel)

# visualize model ..hm, not working
linear_reg_model <- 
  ggPredict(temp_model, se = TRUE, interactive = TRUE)
print(linear_reg_model)

# test to see if fewer vars make model turn out differently
temp_model_2 <- lm(data = chem_phys_subset, temperature_C ~ year4 + depth)
summary(temp_model_2)
# taking daynumber away we get 73.92% of variance explained

temp_model_3 <- lm(data = chem_phys_subset, temperature_C ~ daynum + depth)
summary(temp_model_3)
# without year and just daynum we get 74.15% of variance explained
print(summary(temp_model))
```

5. What is the final set of explanatory variables that predict temperature from your multiple regression? How much of the observed variance does this model explain?

> Answer: The final set of explanatory variables determined by our AIC analysis are year, day number and depth (the same variables we started with). The model explains 74.17% of variance. 

6. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#6
# want to see dif between main and interaction
temp_ANCOVA_main <- lm(data = chem_phys_subset, temperature_C ~ depth + lakename)
temp_ANCOVA_interaction <- lm(data = chem_phys_subset, temperature_C ~ depth * lakename)

summary(temp_ANCOVA_main) # 77.73% var explained
summary(temp_ANCOVA_interaction) # 78.57% var explained
```

7. Is there a significant interaction between depth and lakename? How much variance in the temperature observations does this explain?

> Answer: 

8. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#8
temp_depth <- 
  ggplot(chem_phys_subset, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = .5) +
  ylim(0, 35) +
  labs(x = "Depth (m)", y = "Temperature (˚C)") +
  geom_smooth(method = 'lm', se = FALSE)

print(temp_depth)

```

9. Run a mixed effects model to predict dry mass of litter. We already know that nlcdClass and functionalGroup have a significant interaction, so we will specify those two variables as fixed effects with an interaction. We also know that litter mass varies across plot ID, but we are less interested in the actual effect of the plot itself but rather in accounting for the variance among plots. Plot ID will be our random effect.

a. Build and run a mixed effects model.
b. Check the difference between the marginal and conditional R2 of the model. 

```{r}


```

b. continued... How much more variance is explained by adding the random effect to the model? 

> Answer: 2 %

c. Run the same model without the random effect.
d. Run an anova on the two tests. 
```{r}
 

```

d. continued... Is the mixed effects model a better model than the fixed effects model? How do you know?

> Answer: 