---
title: 'Assignment 7: GLMs week 2 (Linear Regression and beyond)'
author: "Cristiana Falvo"
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 25 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory, load the tidyverse, nlme, and piecewiseSEM packages, import the *raw* NTL-LTER raw data file for chemistry/physics, and import the processed litter dataset. You will not work with dates, so no need to format your date columns this time.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
getwd()
library(tidyverse)
library(nlme)
library(piecewiseSEM)
# to visualize linear reg
require(ggiraph)
require(ggiraphExtra)
require(plyr)
# data
chem_phys <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
litter <- read.csv("/Users/cristiana/Documents/Duke/DataAnalytics/Environmental_Data_Analytics_2020/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 
  
  # is this asking, what's the optimal combination of variables (aic..) to predict lake temps in july?

3. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years. (182-212)
* Only the columns: lakename, year4, daynum, depth, temperature_C (select())
* Only complete cases (i.e., remove NAs) (na.omit(data))

4. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 
  # run aic, this will give optimal combo of variables
  # with this optimal set of vars, run a 'multiple regression'

```{r}
#3 get subset of data
chem_phys_subset <- 
  chem_phys %>%
  filter(daynum >= 182 & daynum <= 212) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>% # 10,836
  na.omit()

#4 run AIC to determine best combination of predictor variables
temp_AIC <- lm(data = chem_phys_subset, temperature_C ~ year4 + daynum + depth) 
step(temp_AIC) 

temp_model <- lm(data = chem_phys_subset, temperature_C ~ year4 + daynum + depth)
summary(temp_model) 
# is this model the multiple regression?
# is there a way to visualize it?
  # can visualize individual relatinships (switch between year, daynum, depth - depth is only one with noteable correlation)
linear_reg_rel <- 
  ggplot(chem_phys_subset, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = .5) +
  geom_smooth(method = lm)
print(linear_reg_rel)

```

5. What is the final set of explanatory variables that predict temperature from your multiple regression? How much of the observed variance does this model explain?

> Answer: The final set of explanatory variables determined by our AIC analysis are year, day number and depth (the same variables we started with). The model explains 74.17% of variance. 

6. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#6
# want to see dif between main and interaction
# temp_ANCOVA_main <- lm(data = chem_phys_subset, temperature_C ~ depth + lakename)
temp_ANCOVA_interaction <- lm(data = chem_phys_subset, temperature_C ~ depth * lakename)

# summary(temp_ANCOVA_main) # 77.73% var explained
summary(temp_ANCOVA_interaction) # 78.57% var explained
```

7. Is there a significant interaction between depth and lakename? How much variance in the temperature observations does this explain?

> Answer: Our results show that there is a significant interaction between depth and lakename, meaning that depth varies significantly depending on which lake we're sampling. Our interaction effects model explains 78.57% of variance. 

8. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#8
temp_depth <- 
  ggplot(chem_phys_subset, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = .5) +
  ylim(0, 35) +
  labs(x = "Depth (m)", y = "Temperature (˚C)") +
  geom_smooth(method = 'lm', se = FALSE)

print(temp_depth)

```

9. Run a mixed effects model to predict dry mass of litter. We already know that nlcdClass and functionalGroup have a significant interaction, so we will specify those two variables as fixed effects with an interaction. We also know that litter mass varies across plot ID, but we are less interested in the actual effect of the plot itself but rather in accounting for the variance among plots. Plot ID will be our random effect.
# litter dataset - looking to predict drymass of litter
  # fixed effects = nlcdClass, functionalGroup (there is interaction between these)
  # random effects = plotID

a. Build and run a mixed effects model.
b. Check the difference between the marginal and conditional R2 of the model. 

```{r}
LitterDrymass_mixed <- 
  lme(data = litter, dryMass ~ nlcdClass * functionalGroup, random = ~1|plotID)

summary(LitterDrymass_mixed) #
rsq <- rsquared(LitterDrymass_mixed)
rsq

R2_diff <- as.numeric(rsq$Marginal - rsq$Conditional)
R2_diff

```

b. continued... How much more variance is explained by adding the random effect to the model? 

> Answer: ~2.13 %

c. Run the same model without the random effect.
d. Run an anova on the two tests. 
```{r}
LitterDrymass_fixed <- 
  lm(data = litter, dryMass ~ nlcdClass * functionalGroup) 
# we were told that the two variables have interaction, thus we multiply


summary(LitterDrymass_fixed)
summary(LitterDrymass_mixed)
# lme for mixed effects model, lm for fixed effects model

comparison_FixedMixed <- anova(LitterDrymass_mixed, LitterDrymass_fixed)
comparison_FixedMixed
```

d. continued... Is the mixed effects model a better model than the fixed effects model? How do you know?

> Answer: Our anova results tell us that the mixed effects model has a lower AIC value, suggesting it is a simpler/more efficient model. We can also see that the mixed and fixed models are significantly different from one another, given our small p-value, so we can conclude that the mixed effects model does a better job at predicting drymass than the fixed effects model.